---
title: "DischargeDistributions"
author: "Camilla Negri"
date: "16/11/2022"
output:
---
## Discharge Distributions

A notebook to fit monthly Q (discharge) distributions to daily data in order to take advantage of the high resolution dataset, hopefully this means the discharge predictions of the BBN can become more accurate.
Purpose is bootstrapping daily Q and fit monthly lognormal distributions.

# Loading the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(plyr)
library(dplyr)  
library(lubridate)
library(lme4)
library(tidyverse)
library(fitdistrplus)

hydrology<-read.csv("data/Ballycanew_discharge.csv")
```

Now lets load the daily data 1 October 2009 - 31 December 2016
Qcumecs = total daily discharge [cubic meters]

# Loading summary data
```{r}
names(hydrology)[names(hydrology) == 'ï..Date'] <- 'Date'
summary(hydrology)
quantile(hydrology$Qcumecs, na.rm=T,probs=c(0.05,0.25,0.5,0.75,0.95,1))
```

# Parsing the data by month
```{r parsing_data}
hydrology$Date<- parse_date_time(hydrology$Date, "d/m/Y")

hydrology$day <- yday(hydrology$Date)
hydrology$months <- month(hydrology$Date)
hydrology$year <- year(hydrology$Date)

# assign months
hydrology$month<-NA # new blank column
hydrology$month[hydrology$months %in% c(1)]<-"January"
hydrology$month[hydrology$months %in% c(2)]<-"February"
hydrology$month[hydrology$months %in% c(3)]<-"March"
hydrology$month[hydrology$months %in% c(4)]<-"April"
hydrology$month[hydrology$months %in% c(5)]<-"May"
hydrology$month[hydrology$months %in% c(6)]<-"June"
hydrology$month[hydrology$months %in% c(7)]<-"July"
hydrology$month[hydrology$months %in% c(8)]<-"August"
hydrology$month[hydrology$months %in% c(9)]<-"September"
hydrology$month[hydrology$months %in% c(10)]<-"October"
hydrology$month[hydrology$months %in% c(11)]<-"November"
hydrology$month[hydrology$months %in% c(12)]<-"December"

# %in% value matching = returns a vector of the positions of (first) matches of its first argument in its second

hydrology$month <- factor(hydrology$month, levels=month.name) 
# encodes a vector as a factor
apply(hydrology, 2, function(x) any(is.na(x)))
```

Now the daily hydrology dataset can be filtered by month
```{r}
hydrology_Qcumecs<-
  hydrology%>%
  dplyr::select(month, Qcumecs)

# check if any NAs 
apply(hydrology_Qcumecs, 2, function(x) any(is.na(x)))
hydrology_Qcumecs
```

Now months as columns, Q cumecs as values 
```{r}
hydrology_Qcumecs<-hydrology_Qcumecs %>%
  group_by(month) %>%
  mutate(row = row_number()) %>%
  tidyr::pivot_wider(names_from = month, values_from = Qcumecs)%>%
  dplyr::select(-row)

hydrology_Qcumecs
```

This gets me a tibble, which plotdist really does not like, using `pull()` might be the solution
```{r}
as.data.frame(hydrology_Qcumecs)
daily_January<-pull(hydrology_Qcumecs,January)
```

# distribution of daily Q in January
```{r}
plotdist(c(na.exclude(daily_January)), histo = TRUE, demp = TRUE) # using c(na.exclude()) to get rid of "Error in density.default(data) : 'x' contains missing values" and get CDF
```
which definitely looks like a Lognormal (as expected).

# Use `fitdist` from `fitdistrplus` to fit the Lognormal
```{r}
Jan_lognormal <- fitdist(daily_January[!is.na(daily_January)], "lnorm") # both !is.na and na.exclude still produce the warning 
summary(Jan_lognormal) # gives std errors too
```
Jan_lognormal$estimate[1] #extracts the mean
Jan_lognormal$estimate[2] #extracts the sd

# Draw a random lognormal with `rlnorm`
```{r}
set.seed(666)
sum(rlnorm(n = 30, mean = Jan_lognormal$estimate[1], sd = Jan_lognormal$estimate[2])) # sum of the randomly drawn lognormal, this means I have a total monthly Q 
```

# Generate 1000 lognormals, sum the 30 days in each
```{r}
monthlyQ<-lapply(1:1000, function(i) sum(rlnorm(n = 30, mean = Jan_lognormal$estimate[1], sd = Jan_lognormal$estimate[2])))
```

# This would get mean and sd, no need to run every time
```{r}
# lapply(1:1000, function(i) {
#     x <- rlnorm(n = 30, mean = Jan_lognormal$estimate[1], sd = Jan_lognormal$estimate[2])
#     c(mean=mean(x), sd=sd(x))
# })
```

# And this would return a matrix of means and sds, no need to run every time
```{r}
# do.call(rbind, lapply(1:1000, function(i) {
#     x <- rlnorm(n = 30, mean = Jan_lognormal$estimate[1], sd = Jan_lognormal$estimate[2])
#     c(mean=mean(x), sd=sd(x))
# }))
```

# distribution of monthly Q in January
```{r}
vectorQ<-unlist(monthlyQ) #you have to unlist the list so it can be plotted
plotdist(vectorQ, histo = TRUE, demp = TRUE)
```

# Use `fitdist` from `fitdistrplus` to fit the Lognormal
```{r}
Jan_monthly_lognormal <- fitdist(vectorQ, "lnorm") # natural log
summary(Jan_monthly_lognormal)
```

-std is asymmetric in linear space but symmetric in logspace 
-natural log is used by `fitdist`

To transform lower and upper limits in linear space, for example: 2.71**(13.8-0.17), and 2.71**(13.8+0.17).

Ok, since this worked with one column of my df, I am going to try to apply the same procedure to all the columns, in order to get a fitted distribution for each month

# Write function to apply to all hydrology_Qcumecs columns, so to each month
```{r}
monthly_lognormal_estimates <- function(MYCOLUMN) {
  daily_lognormal<-fitdist(MYCOLUMN[!is.na(MYCOLUMN)], "lnorm") #fit lnorm to daily data
  monthlyQ<-lapply(1:1000, function(i) sum(rlnorm(n = 30, mean = daily_lognormal$estimate[1], sd = daily_lognormal$estimate[2]),na.rm = T)) # generate 1000 lnorm with the fitted params above, then sum the 30 datapoints in each lnorm so to cumulate days into months
  vectorQ<-unlist(monthlyQ) # convert into vec which is what is needed in fitdist
  monthly_lognormal<-fitdist(vectorQ[!is.na(vectorQ)], "lnorm") # fit monthly lnorm
  return(monthly_lognormal$estimate) # return parameters
} 
```

# Test this function is working with the already extracted January dataset
```{r warning=F}
monthly_lognormal_estimates(daily_January)
```
Looks like it works!

# Apply function to all columns
```{r warning=F}
monthly_lnorm_params<-apply(hydrology_Qcumecs, 2, monthly_lognormal_estimates)
```

# make results more readable with a df
```{r}
# transpose and convert to dataframe
monthly_lnorm_params<-as.data.frame(t(monthly_lnorm_params))

# convert to base 10
monthly_lnorm_params$meanlog10<-monthly_lnorm_params$meanlog/log(10)
monthly_lnorm_params$sdlog10<-monthly_lnorm_params$sdlog/log(10)

monthly_lnorm_params
```

```{r}
write.csv(monthly_lnorm_params,"data/monthly_discharge_distributions.csv")
```


# sample for each month
We can sample from the fitted lognormal 
```{r}
sampled_observedQ<-apply(monthly_lnorm_params, 1, function (i) rlnorm(10000, meanlog = monthly_lnorm_params$meanlog, sdlog = monthly_lnorm_params$sdlog))

sampled_observedQ<-as.data.frame(sampled_observedQ)
head(sampled_observedQ)
```

# Save csv
```{r}
write.csv(sampled_observedQ,"data/monthly_fitted_Q_observations_Ballycanew.csv" )
```

