---
title: "Bayesian network modelling of phosphorus pollution in agricultural catchments"
author: "Camilla Negri"
date: "10/10/2022"

---

## Introduction
This is a notebook aimed at analyzing the results from the two BBNs developed in this studies. The code for the figures is in a separate notebook.

## Aim 
The BBN results extracted from GeNie are analyzed here. We  calculate means and medians of the two models (A describing diffuse sources, and B with the addition of point sources in the form of farmyards and septic tanks), as well as the percentage bias.  


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(SHELF)
library(hydroGOF)
```

Load data and model predictions
```{r}
# load observed data 
hydrology<-read.csv("data/Ballycanew_data.csv")

# Read csv of fitted to observed monthly TRP
sampled_observedTRP<-read.csv("data/monthly_fitted_TRP_observations.csv" ) # these are obtained in the ObservedMonthlyP.Rmd file
```

Model predictions are obtained by running the Model(s) in genie and then extracting the values from the node "In stream TRP concentration mg L-1" (the node handle or identifier is "PConcentration")
```{r}
# model predictions
diffuseconcentrations<-read.csv("results/20230920_DiffusePtool_v4.csv", header=T)
names(diffuseconcentrations)[names(diffuseconcentrations) == 'ï..Pconcentration_mgL']<- 'Pconcentration_mgL' 

head(diffuseconcentrations) # this is Model A

global_ptool_conc<-read.csv("results/20230920_Ptool_v4.csv", header=T)
names(global_ptool_conc)[names(global_ptool_conc) == "PconcentrationmgL"]<- 'Pconcentration_mgL' 
#head(global_ptool_conc) # this is Model B

# monthly data
diffuseconcentrations_months<-read.csv("results/20230920_DiffusePtool_v4_months.csv", header=T)
names(diffuseconcentrations_months)[names(diffuseconcentrations_months) == 'ï..January']<- 'January'
#head(diffuseconcentrations_months)

global_ptool_conc_months<-read.csv("results/20230920_Ptool_v4_months.csv", header=T)
names(global_ptool_conc_months)[names(global_ptool_conc_months) == 'ï..January']<- 'January'
#head(global_ptool_conc_months)
```

Replace negative values in the results with 0s
This is a bit redundant because later in the code we censor the data by the limits of detection. 
```{r}
global_ptool_conc[global_ptool_conc < 0] <- 0
global_ptool_conc_months[global_ptool_conc_months < 0] <- 0
```

##Analyze Phohorus concentrations in the stream [Oct 2009-Dec 2016]

# Load observed P concentrations
TRP = Total Reactive Phosphorus [mg/L], measured by Phosphax every 10 minutes, here daily average

```{r}
# rename P column
hydrology$TRP<-hydrology$TRP..mg.l.
summary(hydrology)
quantile(hydrology$TRP)
```

# Parsing data a monthly time step
```{r parsing_data}
hydrology$Date<- parse_date_time(hydrology$Date, "d/m/Y")

hydrology$day <- yday(hydrology$Date)
hydrology$months <- month(hydrology$Date)
hydrology$year <- year(hydrology$Date)

# assign months
hydrology$month<-NA # new blank column
hydrology$month[hydrology$months %in% c(1)]<-"January"
hydrology$month[hydrology$months %in% c(2)]<-"February"
hydrology$month[hydrology$months %in% c(3)]<-"March"
hydrology$month[hydrology$months %in% c(4)]<-"April"
hydrology$month[hydrology$months %in% c(5)]<-"May"
hydrology$month[hydrology$months %in% c(6)]<-"June"
hydrology$month[hydrology$months %in% c(7)]<-"July"
hydrology$month[hydrology$months %in% c(8)]<-"August"
hydrology$month[hydrology$months %in% c(9)]<-"September"
hydrology$month[hydrology$months %in% c(10)]<-"October"
hydrology$month[hydrology$months %in% c(11)]<-"November"
hydrology$month[hydrology$months %in% c(12)]<-"December"

# %in% value matching = returns a vector of the positions of (first) matches of its first argument in its second

hydrology$month <- factor(hydrology$month, levels=month.name) 
# encodes a vector as a factor
head(hydrology)
```

# Plot daily TRP mg/L by month
```{r}
hydrology%>%
  ggplot(aes(x=TRP))+
  geom_histogram()+
  facet_wrap(~month)

hydrology%>%
  ggplot(aes(x=log10(TRP)))+
  geom_histogram()+
  facet_wrap(~month)
```


# Average mean and sd of TRP by month
These can be compared to mean monthly Genie concentrations
```{r}
monthly_TRP<-hydrology%>%
  group_by(month,months)%>%
  filter(TRP>= 0.01 & TRP <= 5)%>%
  summarise(mean_trp=mean(log10(TRP)), # mean and sd of log TRP
            sd_trp=sd(log10(TRP)))%>%
  mutate(low_trp=mean_trp-sd_trp, # lower and upper limits (1 standard dev.)
         upper_trp=mean_trp+sd_trp)%>%
  dplyr::select(-sd_trp)%>%
  mutate(mean_trp=10**mean_trp, # convert back to linear space so mg L
         low_trp=10**low_trp,
         upper_trp=10**upper_trp)

monthly_TRP
```

# monthly median
```{r}
monthly_TRP_median<-hydrology%>%
  group_by(month,months)%>%
  filter(TRP>= 0.01 & TRP <= 5)%>%
summarise(median_trp=median(TRP,na.rm=T))
monthly_TRP_median
```

# Find mean, lower, upper limits for overall instead of each month
```{r}
overall_monthly_TRP<-hydrology%>%
  filter(TRP>= 0.01 & TRP <= 5)%>%
  summarise(mean_trp=mean(log10(TRP)), # mean and sd of log TRP
            sd_trp=sd(log10(TRP)))%>%
  mutate(low_trp=mean_trp-sd_trp, # lower and upper limits (1 standard dev.)
         upper_trp=mean_trp+sd_trp)%>%
  dplyr::select(-sd_trp)%>%
  mutate(mean_trp=10**mean_trp, # convert back to linear space so mg L
         low_trp=10**low_trp,
         upper_trp=10**upper_trp)

overall_monthly_TRP
```

```{r}
overall_TRP_observations<-hydrology%>%
  filter(TRP>= 0.01 & TRP <= 5)%>%
    summarise(median_trp=median(TRP,na.rm=T),
            quant5th=quantile(TRP,0.05),
            quant25th=quantile(TRP,0.25),
            quant75th=quantile(TRP,0.75))
overall_TRP_observations
```


# Summarize TRP 
```{r describing_data}
hydrology1<-dplyr::group_by(hydrology,year,month)

hydrology_summary<- dplyr::summarize(hydrology1,countTRP=sum(!is.na(TRP)), TRPtotal=sum(TRP,na.rm=T), TRPmean=TRPtotal/countTRP)

Pconc1<-dplyr::group_by(hydrology,year,month)

# to obtain means and quantiles for TRP 
Pconc_summary<- dplyr::summarize(hydrology1, countTRP=sum(!is.na(TRP)),TRPmean=mean(TRP,na.rm=T),TRPtotal=sum(TRP,na.rm=T),TRP_5quant=quantile(TRP, probs =0.05), TRP_25quant=quantile(TRP, probs = 0.25), TRPmedian=median(TRP,na.rm=TRUE), TRP_75quant=quantile(TRP, probs = 0.75))
```

# Summary of TRP per month 
```{r, monthly summary}
hydrology2<-dplyr::group_by(hydrology_summary,month)

hydrology_months_summary<- dplyr::summarize(hydrology2,         countTRPtotal=sum(!is.na(TRPtotal)), TRPmean=mean(TRPtotal,na.rm=T))

Pconc2<-dplyr::group_by(Pconc_summary,month)

Pconc_months_summary<- dplyr::summarize(Pconc2, countTRPtotal=sum(!is.na(TRPtotal)), TRPmean=mean(TRPmean,na.rm=T),                              meanTRP_5quant=mean(TRP_5quant, na.rm=T), meanTRP_25quant=mean(TRP_25quant, na.rm=T), meanTRPmedian=mean(TRPmedian,na.rm=T), meanTRP_75quant=mean(TRP_75quant, na.rm=T)) 
```


# Get TRP values based on lots of probabilities, not just a few quantiles, get all quantiles from 0 to 100
```{r get quantiles}
TRP_probs<-seq(0, 1, 0.01)

TRP_values<-Pconc2 %>%
  dplyr::summarize(quant = quantile(TRPmean, probs=TRP_probs)) %>%
  mutate(qtile = TRP_probs) %>%
  group_by(qtile) %>%
  dplyr::summarize(ave = mean(quant)) # group by quantiles and get avg
      
TRP_values
```

Values are in the ave column, probabilities are in qtile column. These will be used to get a distribution with the `SHELF` package.

# Distribution fitted by averaging monthly quantiles of TRP [mg/L] using 1212 datapoints
```{r}
myTRPfit<-fitdist(vals=TRP_values$ave, probs = TRP_values$qtile, lower = 0)
plotfit(myTRPfit)
```

Sample from the elicited TRP distribution
```{r}
sampled_observedTRP<-sampleFit(myTRPfit, 10000) 
lognormal_observed_TRP<-sampled_observedTRP[1:10000, 4] #subset only the column we're interested in
```

`lognormal_observed_TRP` is the distribution that will be used to calculate the percentage bias. 

#############################################################################################################################

## Analyze Diffuse Ptool (Model A)
Summarize simulations results
```{r}
names(diffuseconcentrations)[names(diffuseconcentrations) == 'ï..Pconcentration_mgL']<- 'Pconcentration_mgL'

summary(diffuseconcentrations)
quantile(diffuseconcentrations$Pconcentration_mgL, probs=c(0.05, 0.25, 0.5, 0.75)) # these are uncensored predictions
```

# Get mean, lower and upper limit (filtered by LOD)
```{r}
overall_monthly_TRP_diffuseBBN<-diffuseconcentrations%>%
  filter(Pconcentration_mgL>= 0.01 & Pconcentration_mgL <= 5)%>%
  summarise(mean_trp=mean(log10(Pconcentration_mgL)), # mean and sd of log TRP
            sd_trp=sd(log10(Pconcentration_mgL)))%>%
  mutate(low_trp=mean_trp-sd_trp, # lower and upper limits (1 standard dev.)
         upper_trp=mean_trp+sd_trp)%>%
  dplyr::select(-sd_trp)%>%
  mutate(mean_trp=10**mean_trp, # convert back to linear space so mg L
         low_trp=10**low_trp,
         upper_trp=10**upper_trp)

overall_monthly_TRP_diffuseBBN
```

```{r}
quantiles_diffuse_TRP_BBN<-diffuseconcentrations%>%
  filter(Pconcentration_mgL >= 0.01 & Pconcentration_mgL <= 5)%>%
  summarise(median_trp=median(Pconcentration_mgL,na.rm=T),
            quant5th=quantile(Pconcentration_mgL,0.05),
            quant25th=quantile(Pconcentration_mgL,0.25),
            quant75th=quantile(Pconcentration_mgL,0.75))

quantiles_diffuse_TRP_BBN
```
 
## Analyze Ptool point and diffuse (Model B)
This adds point sources: Septic Tanks and Farmyards. 
```{r}
names(global_ptool_conc)[names(global_ptool_conc) == 'ï..PconcentrationmgL'] <- 'Pconcentration_mgL'

summary(global_ptool_conc)
quantile(global_ptool_conc$Pconcentration_mgL, probs=c(0.05, 0.25, 0.5, 0.75)) # these are uncensored predictions
```

# Get mean, lower and upper limit for the global model (filtered by LOD)
```{r}
overall_monthly_TRP_BBN<-global_ptool_conc%>%
  filter(Pconcentration_mgL>= 0.01 & Pconcentration_mgL <= 5)%>%
  summarise(mean_trp=mean(log10(Pconcentration_mgL)), # mean and sd of log TRP
            sd_trp=sd(log10(Pconcentration_mgL)))%>%
  mutate(low_trp=mean_trp-sd_trp, # lower and upper limits (1 standard dev.)
         upper_trp=mean_trp+sd_trp)%>%
  dplyr::select(-sd_trp)%>%
  mutate(mean_trp=10**mean_trp, # convert back to linear space so mg L
         low_trp=10**low_trp,
         upper_trp=10**upper_trp)

overall_monthly_TRP_BBN
```

```{r}
quantiles_overall_TRP_BBN<-global_ptool_conc%>%
  filter(Pconcentration_mgL >= 0.01 & Pconcentration_mgL <= 5)%>%
  summarise(median_trp=median(Pconcentration_mgL,na.rm=T),
            quant5th=quantile(Pconcentration_mgL,0.05),
            quant25th=quantile(Pconcentration_mgL,0.25),
            quant75th=quantile(Pconcentration_mgL,0.75))

quantiles_overall_TRP_BBN
```

#Compare each month's concentrations with fitted to observed TRP concentrations 
```{r}
global_ptool_conc_months1<-global_ptool_conc_months
global_ptool_conc_months1$FittedLogNormal<-lognormal_observed_TRP
```

```{r}
model_labels<-c("January","February","March","April",          "May","June","July","August","September","October","November","December","Fitted to Observed")

global_ptool_conc_months1<-global_ptool_conc_months1%>%
  pivot_longer(c("January","February","March","April","May","June","July","August","September","October","November","December","FittedLogNormal"),values_to="concentrations",names_to="model")
``` 

# Comparison dataframe
one dataframe with models and observations
```{r}
# make dataframe
comparison_df <- data.frame(global_ptool_conc$Pconcentration_mgL, diffuseconcentrations$Pconcentration_mgL,lognormal_observed_TRP)

# simplify column names 
names(comparison_df)[names(comparison_df) == 'global_ptool_conc.Pconcentration_mgL'] <- 'global_ptool'
names(comparison_df)[names(comparison_df) == 'diffuseconcentrations.Pconcentration_mgL'] <- 'diffuse_ptool'
names(comparison_df)[names(comparison_df) == 'lognormal_observed_TRP'] <- 'observed'
```

# Count valid samples
```{r}
count_global<-comparison_df%>%
  filter(global_ptool>= 0.01 & global_ptool <= 5)  %>%
  count() 

count_diffuse<-comparison_df%>%
  filter(diffuse_ptool>= 0.01 & diffuse_ptool <= 5)  %>%
  count()

count_fit_obs<-comparison_df%>%
  filter(observed>= 0.01 & observed <= 5)  %>%
  count()

count_global
count_diffuse
count_fit_obs
```

global 8031
diffuse 7221

## Goodness of fit 
Compare GeNie concentrations of both Models with fitted TRP from observed. This is possible because we previously produced the `lognormal_observed_TRP` distribution.

# Filter out the concentrations that are out of limit detection
```{r}
filtered_global<-comparison_df%>%
  filter(global_ptool>= 0.01 & global_ptool <= 5)%>%
  subset(select = -c(diffuse_ptool, observed))
filtered_diffuse<-comparison_df%>%
  filter(diffuse_ptool>= 0.01 & diffuse_ptool <= 5)%>%
  subset(select = -c(global_ptool, observed))
```


# Load predictions and observations again
```{r upload monthly data}
# monthly data
monthly_concentrations<-read.csv("results/20230920_Ptool_v4_months.csv", header=T)
names(monthly_concentrations)[names(monthly_concentrations) == 'ï..January']<- 'January'

# Read csv of fitted to observed monthly TRP
sampled_observedTRP<-read.csv("data/monthly_fitted_TRP_observations.csv" )
```

filter monthly TRP observations 
```{r}
filtered_monthly_concentrations<-monthly_concentrations

filtered_monthly_concentrations$January <- replace(monthly_concentrations$January, monthly_concentrations$January <= 0.01 | monthly_concentrations$January >= 5, NA)

filtered_monthly_concentrations$February <- replace(monthly_concentrations$February, monthly_concentrations$February <= 0.01 | monthly_concentrations$February >= 5, NA)

filtered_monthly_concentrations$March <- replace(monthly_concentrations$March, monthly_concentrations$March <= 0.01 | monthly_concentrations$March >= 5, NA)

filtered_monthly_concentrations$April <- replace(monthly_concentrations$April, monthly_concentrations$April <= 0.01 | monthly_concentrations$April >= 5, NA)

filtered_monthly_concentrations$May <- replace(monthly_concentrations$May, monthly_concentrations$May <= 0.01 | monthly_concentrations$May >= 5, NA)

filtered_monthly_concentrations$June <- replace(monthly_concentrations$June, monthly_concentrations$June <= 0.01 | monthly_concentrations$June >= 5, NA)

filtered_monthly_concentrations$July <- replace(monthly_concentrations$July, monthly_concentrations$July <= 0.01 | monthly_concentrations$July >= 5, NA)

filtered_monthly_concentrations$August <- replace(monthly_concentrations$August, monthly_concentrations$August <= 0.01 | monthly_concentrations$August >= 5, NA)

filtered_monthly_concentrations$September <- replace(monthly_concentrations$September, monthly_concentrations$September <= 0.01 | monthly_concentrations$September >= 5, NA)

filtered_monthly_concentrations$October <- replace(monthly_concentrations$October, monthly_concentrations$October <= 0.01 | monthly_concentrations$October >= 5, NA)

filtered_monthly_concentrations$November <- replace(monthly_concentrations$November, monthly_concentrations$November <= 0.01 | monthly_concentrations$November >= 5, NA)

filtered_monthly_concentrations$December <- replace(monthly_concentrations$December, monthly_concentrations$December <= 0.01 | monthly_concentrations$December >= 5, NA)

sum(is.na(filtered_monthly_concentrations))
colSums(is.na(filtered_monthly_concentrations))
```

fix sampled_observedTRP before using it and sample the same amount of rows as in filtered_monthly_concentrations
```{r}
sampled_observedTRP = subset(sampled_observedTRP, select = -X)

sampled_observedTRP<-sampled_observedTRP %>%
  relocate(January, February, March, April, May, June, July, August, September, .before = October)
```


# PBIAS
The optimal value of PBIAS is 0.0, with low-magnitude values indicating accurate model simulation. Positive values indicate overestimation bias, whereas negative values indicate model underestimation bias. Value: Percent bias between sim and obs. The result is given in percentage (%).

Number of obs and number of sim needs to be the same, therefore, when sampling the obs from the lognormal fitted to the observations, we need to have counted the number of sim first. 

# PBIAS Model A
```{r}
sim<-filtered_diffuse$diffuse_ptool # filtered GeNie simulations (Model A)
obs<-sample(lognormal_observed_TRP,7229) # observed fitted to lognormal, 7229 is the number of sims
```

```{r}
pbias(sim, obs, na.rm=T) # only diffuse 
```

# PBIAS Model B
```{r}
sim2<-filtered_global$global_ptool # GeNie simulations global model (Model B)
obs2<-sample(lognormal_observed_TRP,7851) # observed fitted to lognormal, 7851 is the number of sims
```

```{r}
pbias(sim2, obs2, na.rm=F) # global model (Model B)
```

# PBIAS of every month - Model B
```{r}
head(filtered_monthly_concentrations)
head(sampled_observedTRP)
head(diffuseconcentrations_months)
```

```{r}
for(i in 1:ncol(filtered_monthly_concentrations)) { # for-loop over columns
  sim_months<- filtered_monthly_concentrations
  obs_months<-sampled_observedTRP # observed fitted to lognormal
  print(pbias(sim_months[ , i], obs_months[ , i], na.rm=T))
  }
```

# PBIAS of every month - Model A 
(filter the diffuse concentrations first)

```{r}
filtered_diffuse_monthly<-diffuseconcentrations_months

filtered_diffuse_monthly$January <- replace(diffuseconcentrations_months$January, diffuseconcentrations_months$January <= 0.01 | diffuseconcentrations_months$January >= 5, NA)

filtered_diffuse_monthly$February <- replace(diffuseconcentrations_months$February, diffuseconcentrations_months$February <= 0.01 | diffuseconcentrations_months$February >= 5, NA)

filtered_diffuse_monthly$March <- replace(diffuseconcentrations_months$March, diffuseconcentrations_months$March <= 0.01 | diffuseconcentrations_months$March >= 5, NA)

filtered_diffuse_monthly$April <- replace(diffuseconcentrations_months$April, diffuseconcentrations_months$April <= 0.01 | diffuseconcentrations_months$April >= 5, NA)

filtered_diffuse_monthly$May <- replace(diffuseconcentrations_months$May, diffuseconcentrations_months$May <= 0.01 | diffuseconcentrations_months$May >= 5, NA)

filtered_diffuse_monthly$June <- replace(diffuseconcentrations_months$June, diffuseconcentrations_months$June <= 0.01 | diffuseconcentrations_months$June >= 5, NA)

filtered_diffuse_monthly$July <- replace(diffuseconcentrations_months$July, diffuseconcentrations_months$July <= 0.01 | diffuseconcentrations_months$July >= 5, NA)

filtered_diffuse_monthly$August <- replace(diffuseconcentrations_months$August, diffuseconcentrations_months$August <= 0.01 | diffuseconcentrations_months$August >= 5, NA)

filtered_diffuse_monthly$September <- replace(diffuseconcentrations_months$September, diffuseconcentrations_months$September <= 0.01 | diffuseconcentrations_months$September >= 5, NA)

filtered_diffuse_monthly$October <- replace(diffuseconcentrations_months$October, diffuseconcentrations_months$October <= 0.01 | diffuseconcentrations_months$October >= 5, NA)

filtered_diffuse_monthly$November <- replace(diffuseconcentrations_months$November, diffuseconcentrations_months$November <= 0.01 | diffuseconcentrations_months$November >= 5, NA)

filtered_diffuse_monthly$December <- replace(diffuseconcentrations_months$December, diffuseconcentrations_months$December <= 0.01 | diffuseconcentrations_months$December >= 5, NA)

sum(is.na(filtered_diffuse_monthly))
colSums(is.na(filtered_diffuse_monthly))
```

```{r}
for(i in 1:ncol(diffuseconcentrations_months)) { # for-loop over columns
  sim_months2<- filtered_diffuse_monthly # diffuse model months
  obs_months<-sampled_observedTRP # observed fitted to lognormal
  print(pbias(sim_months2[ , i], obs_months[ , i], na.rm=T))
  }
```



################################################################################################################

## SUMMARY STATISTICS

Recall that the csv of fitted to observed monthly TRP is called `sampled_observedTRP`, while `global_ptool_conc_months` is the predicted TRP from the BBN
```{r}
# monthly data
global_ptool_conc_months<-read.csv("results/20230920_Ptool_v4_months.csv", header=T)
names(global_ptool_conc_months)[names(global_ptool_conc_months) == 'ï..January']<- 'January'
# Read csv of fitted to observed monthly TRP
sampled_observedTRP<-read.csv("data/monthly_fitted_TRP_observations.csv" )
sampled_observedTRP<-subset(sampled_observedTRP, select = -c(X))
sampled_observedTRP
```


# Make labels and prepare dataframes
```{r}
labels<-c("January","February","March","April",          "May","June","July","August","September","October","November","December")

sampled_observedTRP1<-sampled_observedTRP%>%
  pivot_longer(c("January","February","March","April","May","June","July","August","September","October","November","December"),values_to="concentrations",names_to="month")

global_ptool_conc_months1<-global_ptool_conc_months%>%
  pivot_longer(c("January","February","March","April","May","June","July","August","September","October","November","December"),values_to="concentrations",names_to="month")

diffuseconcentrations_months1<-diffuseconcentrations_months%>%
  pivot_longer(c("January","February","March","April","May","June","July","August","September","October","November","December"),values_to="concentrations",names_to="month")
```

# Average mean and sd of TRP by month in each model (filtered)
Model A
```{r}
monthly_TRP_diffuse<-diffuseconcentrations_months1%>%
  filter(concentrations>= 0.01 & concentrations <= 5)  %>%
  mutate(months=case_when(month=="January"~1,
                          month=="February"~2,
                          month=="March"~3,
                          month=="April"~4,
                          month=="May"~5,
                          month=="June"~6,
                          month=="July"~7,
                          month=="August"~8,
                          month=="September"~9,
                          month=="October"~10,
                          month=="November"~11,
                          month=="December"~12))%>%
  group_by(months, month)%>%
  summarise(mean_trp=mean(log10(concentrations)), sd_trp=sd(log10(concentrations)))%>%
  mutate(low_trp=mean_trp-sd_trp, # lower and upper limits (1 standard dev.)
         upper_trp=mean_trp+sd_trp)%>%
  dplyr::select(-sd_trp)%>%
  mutate(mean_trp=10**mean_trp, # convert back to linear space so mg L
         low_trp=10**low_trp,
         upper_trp=10**upper_trp)

write.csv(monthly_TRP_diffuse,"results/20230920_filtered_DiffusePtool_v4_monthly_means.csv")
```

Model B
```{r}
monthly_TRP_ptool<-global_ptool_conc_months1%>%
  filter(concentrations>= 0.01 & concentrations <= 5)  %>%
  mutate(months=case_when(month=="January"~1,
                          month=="February"~2,
                          month=="March"~3,
                          month=="April"~4,
                          month=="May"~5,
                          month=="June"~6,
                          month=="July"~7,
                          month=="August"~8,
                          month=="September"~9,
                          month=="October"~10,
                          month=="November"~11,
                          month=="December"~12))%>%
  group_by(months, month)%>%
  summarise(mean_trp=mean(log10(concentrations)), 
            sd_trp=sd(log10(concentrations)))%>%
  mutate(low_trp=mean_trp-sd_trp, # lower and upper limits (1 standard dev.)
         upper_trp=mean_trp+sd_trp)%>%
  dplyr::select(-sd_trp)%>%
  mutate(mean_trp=10**mean_trp, # convert back to linear space so mg L
         low_trp=10**low_trp,
         upper_trp=10**upper_trp)

write.csv(monthly_TRP_ptool,"results/20230920_filtered_Ptool_v4_monthly_means.csv")
```

# Monthly median each model (filtered data)
Model A
```{r}
monthly_TRP_diffuse_median<-diffuseconcentrations_months1%>%
  filter(concentrations>= 0.01 & concentrations <= 5)  %>%
  mutate(months=case_when(month=="January"~1,
                          month=="February"~2,
                          month=="March"~3,
                          month=="April"~4,
                          month=="May"~5,
                          month=="June"~6,
                          month=="July"~7,
                          month=="August"~8,
                          month=="September"~9,
                          month=="October"~10,
                          month=="November"~11,
                          month=="December"~12))%>%
  group_by(months, month)%>%
  summarise(median_trp=median(concentrations,na.rm=T))

write.csv(monthly_TRP_diffuse_median,"results/20230920_filtered_DiffusePtool_v4_monthly_median.csv")
```

Model B
```{r}
monthly_TRP_ptool_median<-global_ptool_conc_months1%>%
  filter(concentrations>= 0.01 & concentrations <= 5)  %>%
  mutate(months=case_when(month=="January"~1,
                          month=="February"~2,
                          month=="March"~3,
                          month=="April"~4,
                          month=="May"~5,
                          month=="June"~6,
                          month=="July"~7,
                          month=="August"~8,
                          month=="September"~9,
                          month=="October"~10,
                          month=="November"~11,
                          month=="December"~12))%>%
  group_by(months, month)%>%
  summarise(median_trp=median(concentrations,na.rm=T))


write.csv(monthly_TRP_ptool_median,"results/20230920_filtered_Ptool_v4_monthly_median.csv")
```


## Compare median predictions with fitted to observed (months= columns of both df)

merge two dataframes in one, marking clearly what is observed and what is predicted. Recall that the csv of fitted to observed monthly TRP is called `sampled_observedTRP`, while `global_ptool_conc_months` is the predicted TRP from the BBN.
```{r}
# add a column to identify what is what
sampled_observedTRP$model<-as.factor("sampled.observations")
global_ptool_conc_months$model<-as.factor("predicted.BBN")

# use rbind to merge the two dfs in one
total <- rbind(sampled_observedTRP, global_ptool_conc_months)
#View(total)
```

pivot
```{r}
total1<-total%>%
  pivot_longer(c("January","February","March","April","May","June","July","August","September","October","November","December"),values_to="concentrations",names_to="month")
total1$month<-as.factor(total1$month)
total1
```


# Calculate mean of concentration (not logged, not filtered) in the two groups
```{r}
#use plyr to calculate mean
mu <- ddply(total1, "model", summarise, grp.mean=mean(concentrations))
mu
```

# Calculate median of concentration (not logged, not filtered) in the two groups
```{r}
# use plyr to calculate median
eta <- ddply(total1, "model", summarise, grp.median=median(concentrations))
eta
```


################################################################

## Compare predictions to actual monthly observations (months= columns of both dataframes)

make a dataframe of monthly TRP observations 
```{r}
monthly_observedTRP<-hydrology%>%
  group_by(month)%>%
  mutate(id = row_number())%>%
  dplyr::select(-Date,-TRP..mg.l.,-day,-months,-year)%>%
  pivot_wider(names_from = month, values_from = TRP,id_cols = id)%>%
  dplyr::select(-id)

monthly_observedTRP
```

# Merge two dataframes: observations and predictions
```{r}
# add a column to identify what is what
monthly_observedTRP$model<-as.factor("observations")
global_ptool_conc_months$model<-as.factor("predictions")

# use rbind to merge the two dfs in one
total_months <- rbind(monthly_observedTRP, global_ptool_conc_months)
total_months
nrow(total_months)

diffuseconcentrations_months$model<-as.factor("predictions")
total_diffuse_months <- rbind(monthly_observedTRP, diffuseconcentrations_months)
total_diffuse_months
nrow(total_diffuse_months)
```

```{r}
total_months1<-total_months%>%
  pivot_longer(c("January","February","March","April","May","June","July","August","September","October","November","December"),values_to="concentrations",names_to="month")
total_months1$month<-as.factor(total_months1$month)
total_months1
nrow(total_months1)

total_diffuse_months1<-total_diffuse_months%>%
  pivot_longer(c("January","February","March","April","May","June","July","August","September","October","November","December"),values_to="concentrations",names_to="month")
total_diffuse_months1$month<-as.factor(total_diffuse_months1$month)
total_diffuse_months1
nrow(total_diffuse_months1)
```

# use plyr to calculate mean
```{r}
mu2 <- ddply(total_months1, "model", plyr::summarise, grp.mean=mean(concentrations, na.rm=T))
mu2 #plyr::summarise or summarise is the same(?)
```

# use plyr to calculate median
```{r}
eta2 <- ddply(total_months1, "model", summarise, grp.median=median(concentrations,na.rm=T))
eta2
```

# log10 of concentrations median or mean 
```{r}
mu_log10 <- ddply(total_months1, "model", summarise, grp.mean_log=mean(log10(concentrations), na.rm=T))
eta_log10 <- ddply(total_months1, "model", summarise, grp.median_log=median(log10(concentrations), na.rm=T))

mu_log10
eta_log10
```

```{r}
mu_diffuselog10 <- ddply(total_diffuse_months1, "model", summarise, grp.mean_log=mean(log10(concentrations), na.rm=T))
eta_diffuselog10 <- ddply(total_diffuse_months1, "model", summarise, grp.median_log=median(log10(concentrations), na.rm=T))

mu_diffuselog10
eta_diffuselog10
```
