---
title: "Turbidity Distributions"
author: "Camilla Negri"
date: "28/11/2022"
---

A notebook to fit distributions of monthly Turbidity (NTU) by bootstrapping daily NTU observations at catchment outlet (2009-2016).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(plyr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(tidyverse)
library(fitdistrplus)

turbidity<- read.csv("data/Ballycanew_Turbidity_Sediments.csv")
colnames(turbidity)

names(turbidity)[names(turbidity) == 'ï..Date'] <- 'Date'
```

In this turbidity dataset:

- Turbidity[NTU] daily mean from data measured at catchment outlet. NTU is the standard accepted turbidity unit and it stands for Nephelometric Turbidity Units 	

- Suspended Sediment[mg/l] sediment concentration derived from turbidity with calibration curve (methods are described in Sherriff, S., Rowan, J.S., Melland, A.R., Jordan, P., Fenton, O., Ó hUallacháin, D., 2015. Investigating suspended sediment dynamics in contrasting agricultural catchments using ex situ turbidity-based suspended sediment monitoring. Hydrology and Earth System Sciences 19, 3349–3363. https://doi.org/10.5194/hess-19-3349-2015)

- SuspendedSedimentLoad[g]	sediment load is calculated by multiplying the daily sediment concentration by the discharge

# Parsing data a monthly time step
```{r parsing_data}
turbidity$Date<- parse_date_time(turbidity$Date, "d/m/Y")

turbidity$day <- yday(turbidity$Date)
turbidity$months <- month(turbidity$Date)
turbidity$year <- year(turbidity$Date)

# assign months
turbidity$month<-NA # new blank column
turbidity$month[turbidity$months %in% c(1)]<-"January"
turbidity$month[turbidity$months %in% c(2)]<-"February"
turbidity$month[turbidity$months %in% c(3)]<-"March"
turbidity$month[turbidity$months %in% c(4)]<-"April"
turbidity$month[turbidity$months %in% c(5)]<-"May"
turbidity$month[turbidity$months %in% c(6)]<-"June"
turbidity$month[turbidity$months %in% c(7)]<-"July"
turbidity$month[turbidity$months %in% c(8)]<-"August"
turbidity$month[turbidity$months %in% c(9)]<-"September"
turbidity$month[turbidity$months %in% c(10)]<-"October"
turbidity$month[turbidity$months %in% c(11)]<-"November"
turbidity$month[turbidity$months %in% c(12)]<-"December"

# %in% value matching = returns a vector of the positions of (first) matches of its first argument in its second

turbidity$month <- factor(turbidity$month, levels=month.name) 
# encodes a vector as a factor

summary(turbidity)
```

Now the daily dataset can be filtered by month
```{r}
turbidity1<-
  turbidity%>%
  dplyr::select(month, Turbidity.NTU.)
```

Now months as columns, NTU cumecs as values 
```{r}
turbidity1<-turbidity1 %>%
  group_by(month) %>%
  mutate(row = row_number()) %>%
  tidyr::pivot_wider(names_from = month, values_from = Turbidity.NTU.)%>%
  dplyr::select(-row)
```

This gets me a tibble, which plotdist really does not like, using `pull()` might be the solution
```{r}
as.data.frame(turbidity1)
daily_January<-pull(turbidity1,January)
```

# distribution of daily NTU in January
```{r}
plotdist(c(na.exclude(daily_January)), histo = TRUE, demp = TRUE) # using c(na.exclude()) to get rid of "Error in density.default(data) : 'x' contains missing values" and get CDF
```

Looks like a lognormal, as expected.

# Use `fitdist` from `fitdistrplus` to fit the Lognormal
```{r}
Jan_lognormal <- fitdist(daily_January[!is.na(daily_January)], "lnorm") # both !is.na and na.exclude still produce the warning 
summary(Jan_lognormal) # gives std errors too
```

# set seed
```{r}
set.seed(666)
```

# Draw a random lognormal with `rlnorm`
```{r}
sum(rlnorm(n = 30, mean = Jan_lognormal$estimate[1], sd = Jan_lognormal$estimate[2])) # sum of the randomly drawn lognormal, this means I have a total monthly NTU 
```

# Generate 1000 lognormals, sum the 30 days in each
```{r}
monthlyNTU<-lapply(1:1000, function(i) sum(rlnorm(n = 30, mean = Jan_lognormal$estimate[1], sd = Jan_lognormal$estimate[2])))
```

# distribution of monthly Q in January
```{r}
vectorNTU<-unlist(monthlyNTU) #you have to unlist the list so it can be plotted
plotdist(vectorNTU, histo = TRUE, demp = TRUE)
```

# Use `fitdist` from `fitdistrplus` to fit the Lognormal
```{r}
Jan_monthly_lognormal <- fitdist(vectorNTU, "lnorm") # natural log
summary(Jan_monthly_lognormal)
```

# Write function to apply to all turbidity1 columns, so to each month
```{r}
monthly_lognormal_estimates <- function(MYCOLUMN) {
  daily_lognormal<-fitdist(MYCOLUMN[!is.na(MYCOLUMN)], "lnorm") #fit lnorm to daily data
  monthlyNTU<-lapply(1:1000, function(i) sum(rlnorm(n = 30, mean = daily_lognormal$estimate[1], sd = daily_lognormal$estimate[2]),na.rm = T)) # generate 1000 lnorm with the fitted params above, then sum the 30 datapoints in each lnorm so to cumulate days into months
  vectorNTU<-unlist(monthlyNTU) # convert into vec which is what is needed in fitdist
  monthly_lognormal<-fitdist(vectorNTU[!is.na(vectorNTU)], "lnorm") # fit monthly lnorm
  return(monthly_lognormal$estimate) # return parameters
} 
```

# Apply function to all columns
```{r warning=F}
monthly_lnorm_params<-apply(turbidity1, 2, monthly_lognormal_estimates)
```

# make results more readable with a df
```{r}
# transpose and convert to dataframe
monthly_lnorm_params<-as.data.frame(t(monthly_lnorm_params))

# convert to base 10
monthly_lnorm_params$meanlog10<-monthly_lnorm_params$meanlog/log(10)
monthly_lnorm_params$sdlog10<-monthly_lnorm_params$sdlog/log(10)

monthly_lnorm_params
```

```{r}
write.csv(monthly_lnorm_params,"data/monthly_turbidity_distributions_Ballycanew.csv")
```
